---
title: "Data Distillery Sip: Part III of Data Distillery Dash Episode 02"
description: |
  A short description of the post.
author:
  - name: Michael Mullarkey, PhD (Click to return to my website)
    url: https://mcmullarkey.github.io
date: 05-18-2021
draft: true
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

## What is Data Distillery Sip?

Sometimes, I do a livestream/blog post combo called Data Distillery Dash. I take a research topic at random^[yikes] or dataset I've never seen^[double yikes], wrangle the data, create a predictive model or two, and write up a blog post: all within 2 hours live on Twitch! All the words and code in the [part I post](https://mcmullarkey.github.io/mcm-blog/posts/2021-05-13-data-distillery-dash-episode-02/){target="_blank"} were written during an ~2 hour livestream. If you want to come join the party, I plan on doing this on Thursdays from 3:30-5:30 EST [here.](https://www.twitch.tv/mcmullarkey){target="_blank"} <br>
<br>
Sometimes, I'll want to do extensions on what I accomplished during the Data Distillery Dash, but at a (slightly) more leisurely pace. I've decided to call those posts Data Distillery Sips. This is the second one (see the first one [here](https://mcmullarkey.github.io/mcm-blog/posts/2021-05-17-data-distillery-sip-a-part-ii-of-data-distillery-dash-episode-02/){target="_blank"}), and also serves as a continuing tutorial on how to build a predictive pipeline for psychological interventions.

## Where Did We Leave Off/Why Were We Doing That?

At the end of the last post, we had explained why we needed to build two different predictive models for this particular prediction problem. Long story short, we need to generate predictions as if people had received both psychological interventions, rather than just one. For a more detailed explantion why that is, check out the [previous post.](https://mcmullarkey.github.io/mcm-blog/posts/2021-05-17-data-distillery-sip-a-part-ii-of-data-distillery-dash-episode-02/){target="_blank"} <br>
<br>
We've built both of those models, and we can read them in now.

```{r}

act_model <- read_rds("active_workflow.rds")
ctrl_model <- read_rds("control_workflow.rds")

```

We'll also read in the holdout data we created in [a previous post](https://mcmullarkey.github.io/mcm-blog/posts/2021-05-13-data-distillery-dash-episode-02/){target="_blank"} so we have data to make our predictions on!

```{r}

holdout_ctrl <- read_rds("test_set_control.rds")
holdout_actv <- read_rds("test_set_active.rds")

```




















